---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(badger)
library(knitr)
library(subsets)
set.seed(1)
```

# subsets

<!-- badges: start -->
`r badge_repostatus("WIP")`
`r badge_lifecycle("experimental")`
`r badge_custom("ORCID", "0000-0001-9498-8379", "green",   "https://orcid.org/0000-0001-9498-8379")`
<!-- badges: end -->



subsets: an Easy Grammar of Subsets


&nbsp;

## Motivation
    
    Among programming languages, 'R' has perhaps one of the most
    flexible and comprehensive sub-setting functionality. But with
    flexibility often comes confusion and (apparent) inconsistencies.
    And 'R' is no exception.
    
    This becomes quite apparent when one reads (online) documents such
    as "The R Inferno" by Patrick Burns, and "Frustration: One Year
    With R" by Reece Goding. These documents point out many
    inconsistencies, and sub-setting related inconsistencies make up a
    good portion of these documents.
    
    To my surprise, there is no comprehensive R-package (as far as I
    could see at least) that actually attempts to "fix" the
    subset-related issues laid out in these and other such documents.
    
    Famous subset-related R packages such as 'dplyr' and 'data.table'
    focus almost exclusive on data.frame-like objects, and
    occasionally even add more frustration in some aspects, like being
    not very programmatically friendly.
    
    Thus, this R package was born.
    
    Although this package was somewhat made for people who are new to
    'R' (especially when coming from another programming language),
    and found themselves confused, I trust this package will be useful
    even for those who are quite experienced in 'R'.
    
&nbsp;

## Goal & Properties


The Goal of the 'subsets' package is not to replace the square-brackets operators (`[`, `[[`, `[<-`, and `[[<-`) per-sé, but to provide **alternative** sub-setting methods and functions, to be used in situations where the square-brackets operators are inconvenient. 

These are (hopefully) easier sub-setting methods and functions with the following properties:

  * *Programmatically friendly*:

      * Non-standard evaluation is quite controversial (and for
        good reasons), and therefore completely absent in this R
        package.

      * Name-based arguments instead of position-based arguments.

      * Missing arguments can be filled with `NULL`, instead of
        using dark magic like `base::quote(expr = )`.

      * Functions are pipe-friendly.

  * *Beginner friendly*:

      * No (silent) vector recycling.

      * Extracting and removing subsets uses the same syntax.

      * All functions return a copy of the object, unless stated
        otherwise.

  * *Class consistent*:

      * sub-setting of multi-dimensional objects by specifying
        dimensions (i.e. rows, columns, ...) use `drop = FALSE`.
        So matrix in, matrix out.

      * The functions deliver the same results for data.frames,
        data.tables, tibbles, and tidytables. No longer does one
        have to re-learn the different brackets-based sub-setting
        rules for different types of data.frame-like objects.
        Powered by the subclass agnostic 'C'-code from 'collapse'
        and 'data.table'.

  * *Explicit copy semantics*:

      * Sub-set operations that change its memory
        allocations, always return a modified copy of the object.

      * For sub-set operations that just change values in-place
        (similar to the [<- and [[<- methods) the user can choose
        a method that modifies the object by *reference*, or choose
        a method that returns a *deep copy*.

  * *Careful handling of names and other attributes*:

      * Sub-setting an object by index names returns ALL indices
        with that name, not just the first.

      * Data.frame-like objects (see supported classes below) are
        forced to have unique column names.

      * Attributes of data.frame-like objects (see supported
        classes below) are always preserved when sub-setting.

      * For other object types, the user can specify whether to
        preserve Attributes, or use R's `[` attribute behaviour
        (i.e. drop most attributes). This is to ensure
        compatibility with R-packages that create their own
        attribute behaviour for sub-setting.

  * *Support a wide variety of S3 classes*:

      * Support atomic objects (vectors, matrices,
        arrays).
        
      * Support factors.

      * Support lists.

      * Support the following data.frame-like objects:
        data.frame, data.table, tibble, and tidytable class, and
        objects derived from these classes.

      * Support for the column selection sub-setting used in
        ggplot2's `aes()` function.

      * Support for sub-setting characters in a single string.

      * Since the main functions are S3 functions, other packages
        may add functionality for additional classes.

  * *Concise function and argument names*.

  * *Performance aware*:
    Despite the many checks performed, the functions are kept
    reasonably speedy, through the use of the 'Rcpp', 'collapse',
    and 'data.table' R-packages.
    Most of the heavy lifting in this package is done by the 'collapse' package.

&nbsp;

## Methods and Functions

The main focus is on the following generic S3 methods:

  * `sb_x()`: method to extract, exchange, or duplicate indices.

  * `sb_rm()`: method to remove indices.
  
  * `sb_set()`: method to modify
    (transform or replace values)
    subsets of an object by **reference**.

  * `sb_mod()`: method to return a **copy**
     of an object with modified
     (transformed or replaced values)
     subsets.
    
  * `sb_coe()`: coerce and transform a whole object,
    or a recursive subset of an object.

  * `sb_before()`, `sb_after()`: methods to insert new values before or
    after an index along a dimension of an object.
  
  * `sb_rec()`: not actually a method,
    but a function that can be combined with the above methods,
    for recursive sub-setting operations.


Beside these generic S3 methods, additional specialized
sub-setting functions are provided:

  * `aes_pro()`: programmatically friendly and stable version of
    the `ggplot2::aes()` function.
    
  * `sb_str()`: extract or replace a subset of characters of a single string
    (each single character is treated as a single element).

  * `sb_a()`: extract multiple attributes from an object.


And finally, a couple of helper functions for creating ranges,
sequences and indices (often needed in sub-setting) are provided:

  * `seq_rec()`: Recursive sequence generator
    (for example to generate a Fibonacci sequence)

  * `seq_names()`: create a range of indices from a specified
    starting and ending name.
    
  * `sub2coord()`, `coord2ind()`: Convert subscripts
    (array indices)
    to coordinates,
    coordinates to flat indices, and vice-versa.

&nbsp;

## Help pages

For an explanation of the classes, and how each class is treated by 'subsets',
see `subsets_classes`.

For an explanation of the common indexing arguments in the generic methods,
see `subsets_indx_args`.

&nbsp;

## Installing & Loading

One can install 'subsets' from GitHub like so:

```{r eval = FALSE, echo = TRUE}
remotes::install_github("https://github.com/tony-aw/subsets")
```

Special care has been taken to make sure the function names are clear, and that the function names are unlikely to conflict with core R, the recommended R packages, the rstudioapi package, or major packages from the fastverse. So one can attach the package - thus exposing its functions to the namespace - using:

```{r eval = FALSE, echo = TRUE}
library(subsets)
```


Should the user wish to expose specific functions from 'subsets' ONLY within a specific environment, like only within a specific function, one can use the following:


```{r eval = FALSE, echo = TRUE}
tinycodet::import_LL("subsets", selection = ... )
```

&nbsp;

## Benchmarks

Due to the many checks and conversions performed by the `subsets::` functions, to make sub-setting more programmatically and beginner friendly,
the functions are almost necessarily slower than base R's `[`-like operators.

However, a considerable effort was made to keep the speed loss to a minimum.
The exact speed loss depends on the situation. For example, the `subsets::` functions are about as fast as base R when sub-setting using names, but base R is faster when sub-setting matrices using only numbers. But sub-setting data.frames are actually faster with the `subsets::` functions than in base R.

Below are some benchmarks to give one an idea of the speed loss. These are just examples; speed is determined by a great number of factors.

&nbsp;

### Matrix

```{r eval=FALSE, echo=TRUE}
x <- matrix(seq_len(1000*900), ncol = 900)
colnames(x) <- sample(c(letters, NA), 900, TRUE)
bm.matrix <- rbenchmark::benchmark(
  "subsets" = sb_x.matrix(x, 1:100, c("a", "a")),
  "base R" = x[1:100, lapply(c("a", "a"), \(i) which(colnames(x) == i)) |> unlist(), drop = FALSE],
  replications = 1e5,
  order = NULL
)
print(bm.matrix)
```


```{r echo = FALSE, eval = TRUE}
load("performance/bm.matrix.RData")
print(bm.matrix)
```


&nbsp;

### Array (4D)

```{r eval=FALSE, echo=TRUE}
x.dims <- c(1000, 900, 4)
x <- array(1:prod(x.dims), x.dims)
idx <- list(1:100, c(TRUE, TRUE, TRUE, FALSE))
dims <- c(1, 3)
bm.array <- rbenchmark::benchmark(
  "subsets" = sb_x.array(x, idx, dims),
  "base R + abind" = abind::asub(x, idx, dims),
  replications = 1e4,
  order = NULL
)
print(bm.array)
```

```{r echo = FALSE, eval = TRUE}
load("performance/bm.array.RData")
print(bm.array)
```



&nbsp;

### Data.frame

```{r eval=FALSE, echo=TRUE}

n <- 1e6
x <- data.frame(
  a = seq_len(n),
  b = sample(letters, size = n, replace = TRUE),
  c = seq_len(n) * -1,
  d = sample(rev(letters), size = n, replace = TRUE)
)
colsel <- rep("a", 4)
bm.df <- rbenchmark::benchmark(
  "subsets" = sb_x.data.frame(x, 1:1000, colsel),
  "base R" = x[1:1000, match(colsel, names(x))],
  replications = 1e4,
  order = NULL
)
print(bm.df)

```

```{r echo = FALSE, eval = TRUE}
load("performance/bm.df.RData")
print(bm.df)
```

&nbsp;


## See Also

'subsets' relies on the 'Rcpp', 'collapse' and 'data.table'
R-packages to ensure an acceptable performance of its functions
despite the many checks that these functions perform. I also
recommend using these packages for other sub-setting and data
wrangling functionalities.
Besides these package, the following R packages work very nicely
together with 'subsets':

  * 'stringi':
    THE R package for fast and concise string manipulation - an
    essential part of any programming language.

  * 'abind':
    Provides binding arrays along an arbitrary dimension.

  * 'tinycodet':
    Helps the user with their coding etiquette. Focuses on 4
    aspects: (1) safe functionalities, (2) an import system that
    combines benefits of using without attaching and attaching a
    package, (3) extending the capabilities of the aforementioned
    'stringi' package, (4) functions to reduce repetitive code.
    
&nbsp;


## Changelog (EXPERIMENTAL VERSIONS)

 - 31 October 2023: Very first GitHub upload. Also: Halloween!
 - 12 November 2023: Optimised the code more. Update the benchmarks with more representative sized objects. Merged `sb_rp()` and `sb_tf()` into `sb_mod()`, and added `sb_set()`. Placed explanations of the common indexing arguments in a separate help page for a better overview (and less typing for me). Replaced the `.attr` argument with the `rat` argument in `sb_x()` and `sb_rm()`. Added the `sub2ind` and `ind2sub` functions.
 - 18 November 2023: Fixed several bugs in the `sb_set()` method. Added the `sb_coe()` method. Added considerably more tests. Expanded the documentation a bit. Expanded the `sub2ind` - functions into 4 more comprehensive functions. The `sb_str()` function now also has a replacement method.
 - 22 November 2023: Made the arguments in the `sub2ind` - functions more consistent, and also expanded their documentation a bit. Added an expanded explanation of the classes for argument `x` in a separate help file for a better overview (and less typing for me). Added more badges. Improved the tests a bit. Added an `.onAttach()` message function. The `aes_pro()` function now uses formula input rather than character input.
 - 27 November 2023: Added the `rcl` argument to the `sb_*.array()` methods, specifically for 3-dimensional arrays. Produced more tests, and re-organised the tests. Added the `with_pro()` function, which - like `aes_pro()` - is the programmatically friendly version of `with()`, and works with a formula instead of expression. Added reference to 'abind' in bibliography of the `sb_before()` and `sb_after()` methods. Normalized the order of the classes in the documentation of all the generic methods of 'subsets'. Added the `n()` function, as short-hand for `list()`, and also functions as the nested version of `c()`. Added an additional help page explaining Non-Standard Evaluation (NSE); 'subsets' avoids the usage of NSE whenever possible.
 

&nbsp;
