---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(badger)
library(knitr)
library(subsets)
set.seed(1)
```

# subsets

<!-- badges: start -->
`r badge_repostatus("WIP")`
`r badge_lifecycle("experimental")`
`r badge_custom("ORCID", "0000-0001-9498-8379", "green",   "https://orcid.org/0000-0001-9498-8379")`
<!-- badges: end -->



subsets: an Easy Grammar of Subsets


&nbsp;

## Motivation
    
    Among programming languages, 'R' has perhaps one of the most
    flexible and comprehensive sub-setting functionality. But with
    flexibility often comes confusion and (apparent) inconsistencies.
    And R is no exception.
    
    This becomes quite apparent when one reads (online) documents such
    as "The R Inferno" by Patrick Burns, and "Frustration: One Year
    With R" by Reece Goding. These document point out many
    inconsistencies, and sub-setting related inconsistencies make up a
    good portion of these documents.
    
    To my surprise, there is no comprehensive R package (as far as I
    could see at least) that actually attempts to "fix" the
    subset-related issues laid out in these and other such documents.
    
    Famous subset-related R packages such as 'dplyr' and 'data.table'
    focus almost exclusive on data.frame-like objects, and
    occasionally even add more frustration in some aspects, like being
    not very programmatically friendly.
    
    Thus, this R package was born.
    
    Although this package was somewhat made for people who are new to
    'R' (especially when coming from another programming language),
    and found themselves confused, I trust this package will be useful
    even for those who are quite experienced in 'R'.
    
&nbsp;

## Properties

The subsets package provides easier subsetting functions with the
following properties:

  * *Programmatically friendly*:

      * Non-standard evaluation is highly controversial (and for
        good reasons!), and therefore completely absent in this R
        package;

      * Name-based instead of position-based arguments;

      * Missing arguments can be filled with `NULL`, instead of
        using dark magic like `base::quote(expr = )`.

      * Functions are pipe-friendly.

  * *Beginner friendly*:

      * No (silent) vector recycling;

      * Extracting and removing subsets uses the same syntax.

  * *Class consistent*:

      * sub-setting of multi-dimensional objects by specifying
        dimensions (i.e. rows, columns, ...) use `drop = FALSE`.
        So matrix in, matrix out.

      * The functions deliver the same results for data.frames,
        data.tables, tibbles, and tidytables. No longer does one
        have to re-learn the different brackets-based sub-setting
        rules for different types of data.frame-like objects.
        Powered by the subclass agnostic 'C'-code from 'collapse'
        and 'data.table'.

      * Smart with sub-setting recursive lists.

  * *Careful handling of name-based indexing*:

      * Sub-setting object by index names returns ALL indices
        with that name, not just the first;

      * Data.frame-like objects are forced to have unique column
        names;

      * Selecting non-existing names always gives an error.

  * *Support a wide variety of data types*:

      * Support vector-like (atomic) objects (vectors, matrices,
        arrays);

      * Support lists;

      * Support factors;

      * Support data.frame-like objects (data.frame, data.table,
        tibble, tidytable, etc.).

  * *Concise function and argument names*.

  * *Special functions*:
    for string subsetting, vectorized recursive list subsetting,
    and even for the column selection subsetting used in
    ggplot2's `aes()` function.

  * *Performance aware*:
    Despite the many checks performed, the functions are kept
    reasonably speedy, through the use of the 'Rcpp', 'collapse',
    and 'data.table' R-packages. See the benchmarks section below for some         examples.

&nbsp;

## Methods and Functions

The main focus is on the following generic S3 methods:

  * `sb_x`: method to extract, exchange, or duplicate indices.

  * `sb_rm`: method to remove indices.

  * `sb_tf`: method to transform values of subsets.

  * `sb_rp`: method to replace values of subsets.

  * `sb_before`, `sb_after`: methods to insert new values before or
    after an index along a dimension of an object.


Beside these generic S3 methods, additional specialized
sub-setting functions are provided:

  * `aes_pro`: programmatically friendly and stable version of
    ggplot2's aesthetic sub-setting function.

  * `sb_str`: subset a single string (each single character is
    treated as a single element).

  * `sb_rec`: recursive sub-setting of lists.

And finally, a couple of helper functions for creating ranges and
sequences (occasionally needed in sub-setting) are provided:

  * `seq_rec`: Generalized recursive sequence generator.

  * `seq_names`: create a range of indices from a specified
    starting and ending name.

&nbsp;

## Installing & Loading

One can install 'subsets' from GitHub like so:

```{r eval = FALSE, echo = TRUE}
remotes::install_github("https://github.com/tony-aw/subsets")
```

Special care has been taken to make sure the function names are clear, and that the function names are unlikely to conflict with core R, the recommended R packages, the rstudioapi package, or major packages from the fastverse. So one can attach the package - thus exposing its functions to the namespace - using:

```{r eval = FALSE, echo = TRUE}
library(subsets)
```


Should the user wish to expose specific functions from 'subsets' ONLY within a specific environment, like only within a specific function, one can use the following:


```{r eval = FALSE, echo = TRUE}
tinycodet::import_LL("subsets", selection = ... )
```

&nbsp;

## Benchmarks

Due to the MANY, MANY checks and conversions performed by the `subsets::` functions, to make sub-setting more programmatically and beginner friendly,
the functions are almost necessarily slower.

However, a considerable effort was made to keep the speed loss to a minimum.
The exact speed loss depends on the situation. For example, the `subsets::` functions are about as fast as base R when sub-setting using names, but base R is much faster when sub-setting matrices using only numbers. But subsetting data.frames are actually faster with the `subsets::` functions than in base R.

Below are some benchmarks to give one an idea of the speed loss. These are just examples; speed is determined by a great number of factors.

```{r}
library(rbenchmark)
library(subsets)
```


```{r eval=FALSE, echo=TRUE}
x <- matrix(1:50000, ncol = 20)
colnames(x) <- c(letters[1:18], "a", NA)
bm.matrix <- benchmark(
  "subsets" = sb_x(x, 1:100, c("a", "a")),
  "base R" = x[1:100, lapply(c("a", "a"), \(i) which(colnames(x) == i)) |> unlist(), drop = FALSE],
  replications = 100000
)
print(bm.matrix)
```


```{r echo = FALSE, eval = TRUE}
load("performance/bm.matrix.RData")
print(bm.matrix)
```


```{r eval=FALSE, echo=TRUE}
x.dims <- c(1000,100,4)
x <- array(1:prod(x.dims), x.dims)
idx <- list(1:100, c(TRUE, TRUE, TRUE, FALSE))
dims <- c(1, 3)
bm.array <- benchmark(
  "subsets" = sb_x(x, idx, dims),
  "base R + abind" = abind::asub(x, idx, dims),
  replications = 10000
)
print(bm.array)
```

```{r echo = FALSE, eval = TRUE}
load("performance/bm.array.RData")
print(bm.array)
```


```{r eval=FALSE, echo=TRUE}
n <- 1e5
x <- data.frame(
  a = seq_len(n),
  b = sample(letters, size = n, replace = TRUE),
  c = seq_len(n) * -1,
  d = sample(rev(letters), size = n, replace = TRUE)
)
bm.df <- benchmark(
  "subsets" = sb_x(x, 1:1000, c("a", "a")),
  "base R" = x[1:1000, lapply(c("a", "a"), \(i) which(colnames(x) == i)) |> unlist(), drop = FALSE],
  replications = 1e4
)
print(bm.df)

```

```{r echo = FALSE, eval = TRUE}
load("performance/bm.df.RData")
print(bm.df)
```

&nbsp;


## See Also

'subsets' relies on the 'Rcpp', 'collapse' and 'data.table'
R-packages to ensure an acceptable performance of its functions
despite the many checks that these functions perform. I also
recommend using these packages for other subsetting and data
wrangling functionalities.
Besides these package, the following R packages work very nicely
together with 'subsets':

  * 'stringi':
    THE R package for fast and concise string manipulation - an
    essential part of any programming language.

  * 'abind':
    Provides binding arrays along an arbitrary dimension.

  * 'tinycodet':
    Helps the user with their coding etiquette. Focuses on 4
    aspects: (1) safe functionalities, (2) an import system that
    combines benefits of using without attaching and attaching a
    package, (3) extending the capabilities of the aforementioned
    'stringi' package, (4) functions to reduce repetitive code.
    
